INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-11-04 chapters/evaluation.tex
evaluation - first part

@@

We implemented the proposed INGRID method as an (in the form of) an MPS plugin that allows users to run (perform, carry out) the construction of an MPS language from an ANTLR v4 grammar.
We used a mix of Java and BaseLanguage in our implementation.
We used the MPS API, which can be called from the BaseLanguage programs too, to generate some of the language elements.
BaseLanguage is needed to use/call the MPS API, e.g. in order to programmatically create new language elements and their aspects.
The plugin uses the ANTLR library~\cite{ref:ANTLR4} to work with ANTLR v4 grammar files (to parse them, etc) and several MPS libraries (that implement the MPS API).
Our complete implementation is available at \todo{url of github repository}.

For the purpose of evaluation, we applied INGRID to several complex mainstream languages, including JSON, JavaScript, and C\#.
We used the version of JavaScript that is standardized as ECMAScript 5.1 --- the specification was released/finalized in 2011 and it is currently the most frequently adopted version.
MPS projects that contain definitions of all these three programming languages are available in the public repository at \todo{url}
In the rest of this section, we discuss our experience with application of INGRID to the selected languages, and then we highlight general observations.

However, before giving details on the experience, we must emphasize that the MPS languages produced by INGRID (the automatically imported languages) are not ready-to-use full-fledged MPS counterparts of the original input languages.
Imported languages contain full structure found in the respective ANTLR grammars and some more aspect definitions that will help users of the languages in creating code.
Every imported language has to be manually tweaked/adjusted by the end user (e.g., in particular/especially its Editor aspect).
The imported language has to be manually adjusted (tweaked, polished) by the end user (developer) and some of its complications have to be resolved manually by a human.
One of our goals was to show that even though our method (plugin) does not create a perfect MPS language, it can be customized very fast into a very usable form.
We manually tweaked/customized the editor and the TextGen aspect only, structure aspect was left in its original form.

For each of the languages, we provide the result (MPS language) exactly in the form as created by our implementation (plugin), and also the adjusted grammars used as input for INGRID, both/all in the repository with implementation.

\paragraph{JSON.}
The first language that we tried was JSON \todo{nejaka reference na specifikaci nebo pripadne gramatiku}.
The adjusted version of the JSON language in MPS was created from the original by the first author in no more than 20 minutes of working on it.
More specifically, the JSON language was ready for use in no more than 20 minutes of minor adjusting.
We chose this language because it does not require us to implement complex aspects, such as type checking or data flows.

\paragraph{JavaScript.}
We have also tried to use the INGRID tool/plugin to import the JavaScript language, which is an example/instance of a widely-used complicated general purpose language.
Importing the JavaScript language~\cite{ref:javascript} and manually adjusting all editors, that needed it, in a less than hour time.

There are already some projects, where a manual port of JavaScript is being done.
One of them is the ECMAScript4MPS~\cite{ref:ECMAScript4MPS} from the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section X (similar projects, related work).
The result of import by INGRID is quite good, comparable to the PE4MPS project.
Our language is, of course, missing a lot of more advanced aspects/features that, for example, can help deriving types of expressions and so on.
These aspects improve the usability by a lot, e.g. when the user starts writing numbers, a number literal is inserted, whereas in our import the user has to first insert the concept representing the number literal and then proceed with writing numbers.
These actions are needed in any MPS language and are expected to be implemented after the import step, \todo{Maybe say that INGRID does not generate these actions/aspects/features yet but it is our future work} as it is not possible to generate them automatically.
However, when it comes to language structure, concept aliases, and auto-completion, we think that INGRID (our method/plugin) did a very good job.
The structure aspect is very similar to the ECMAScript4MPS's one, which was created manually over the course of a large number of hours.
INGRID does that (achieves the same result) fully automatically, and yields very good/nice results.

@@

\paragraph{General observations.}
Most of the languages (that we tried to define in MPS, use/import into MPS) are quite complex when it comes to their structure and the overall syntax variety, so defining them manually would be a very tedious, error-prone and time-consuming process.
\todo{Discuss where INGRID helps the most based on the experience.}
(Our experience/experiments show that) Automated generation of the Structure aspect is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
On the other hand, we found out that manual tweaking/adjusting of the Editor and Textgen aspects (adding line breaks and indentation) is very effective, and takes only a relatively little/short time (much less than we expected).
The imported language can be, in our opinion, adjusted very fast into a more readable form.
In fact/total, the adjusted version of each language was created from the original import by the first author, who spent no more than between 20 to 60 minutes of working on each of them.
The first author spent only a couple of minutes of (manually) adjusting the languages after the automated import.
Our experience with all the languages confirms our hypothesis (that code layout in the Editor aspect can be adjusted very fast, from the basic default single-row layout), and justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical.
It takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language and manually adjusting all editors that needed it, all that in a less than hour time.
\todo{Mozna rict:} However, during our work on this project, we also tried to design some/various automated heuristics, but all that we came up with so far give rather suboptimal/bad results especially when compared to results produced by a human user.

>>>>

We implemented the proposed INGRID method as an MPS plugin that allows users to generate languages from ANTLR v4 grammars.
While most of the plugin is written in Java, small fragments of BaseLanguage code were needed to call the MPS API, e.g. in order to programmatically create new language concepts and their aspects.
The plugin uses the ANTLR library~\cite{ref:ANTLR} for parsing of ANTLR v4 grammar files, and several MPS libraries that implement the MPS API.
Our complete implementation is available at \url{https://github.com/premun/ingrid}.

For the purpose of evaluation, we applied INGRID to several well-established and widely used languages, including JSON, JavaScript, and C\#.
We used the version of JavaScript that is standardized as ECMAScript 5.1~\cite{ref:ECMASCRIPT51} --- the specification was finalized in June 2011, and it is currently the most frequently adopted version.
MPS projects that contain definitions of all three languages are also released in the public repository (\url{https://github.com/premun/ingrid}).
In the rest of this section, we discuss our experience with application of INGRID to these languages, and then we highlight few general observations.

However, first we must emphasize that MPS languages automatically produced by the INGRID method, as defined in this paper, are not ready-to-use full-fledged MPS counterparts of the original input languages.
The structure of such a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects (e.g., the Editor) would have to be manually tweaked (or defined from scratch) by the end user.
INGRID does not yet support advanced features of MPS, such as the aspect responsible for type checking and inference.
For each of the three languages (JSON, JavaScript, and C\#), we provide (1) its MPS definition exactly in the form created by INGRID and (2) the adjusted ANTLR v4 grammar used as input for INGRID, both in the repository that contains also our implementation.

\paragraph{JSON.}
The least amount of manual adjusting after the import into MPS was needed in case of the JSON language~\cite{ref:JSON}, because it is the simplest language from all that we used for our experiments.
Specifically, the first author spent less than 20 minutes in order to get a language that is ready to use.

\paragraph{JavaScript.}
In the case of JavaScript, which is an example of a widely-used complex general purpose language, automated generating of the language definition in MPS from the ANTLR grammar\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} and subsequent manual adjusting was done in less than one hour.

We are aware of other projects that aim to create a manual port of JavaScript into MPS.
For example, there is ECMAScript4MPS~\cite{ref:ECMA4MPS} developed by the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section~\ref{sect:RELATED}.

The main advantage of INGRID over ECMAScript4MPS is that, despite its current limitations, INGRID fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful.
INGRID achieved a very good result especially in the case of language structure, concept aliases, and support for auto-completion.
The Structure aspect generated by INGRID is very similar to that of the ECMAScript4MPS project, which was created manually in a large number of hours.

@@

@@

\paragraph{General observations.}
Most of the languages (that we tried to define in MPS, use/import into MPS) are quite complex when it comes to their structure and the overall syntax variety, so defining them manually would be a very tedious, error-prone and time-consuming process.
\todo{Discuss where INGRID helps the most based on the experience.}
(Our experience/experiments show that) Automated generation of the Structure aspect is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
On the other hand, we found out that manual tweaking/adjusting of the Editor and Textgen aspects (adding line breaks and indentation) is very effective, and takes only a relatively little/short time (much less than we expected).
The imported language can be, in our opinion, adjusted very fast into a more readable form.
The generated MPS language can be customized very fast into a very usable form.
In fact/total, the adjusted version of each language was created from the original import by the first author, who spent no more than between 20 to 60 minutes of working on each of them.
The first author spent only a couple of minutes of (manually) adjusting the languages after the automated import.
Our experience with all the languages confirms our hypothesis (that code layout in the Editor aspect can be adjusted very fast, from the basic default single-row layout), and justifies some of the decisions we have described in Section X (editor aspect) and Section Y (textgen), proving that our approach is quite practical.
It takes a very short time to reach optimal results (because MPS allows doing this very efficiently), as we show on the example of importing the JavaScript language and manually adjusting all editors that needed it, all that in a less than hour time.
\todo{Mozna rict:} However, during our work on this project, we also tried to design some/various automated heuristics, but all that we came up with so far give rather suboptimal/bad results especially when compared to results produced by a human user.

### 2016-11-04 paper.tex
writing section 3
evaluation - first part

@@

@@

% TODO opravit formatovani nasledujicich polozek a vhodne zaradit

\bibitem{ref:ANTLR4} ANTLRv4, \url{https://github.com/antlr/antlr4}

\bibitem{ref:ANTLR4reference} T. Parr. The Definitive ANTLR 4 Reference. The Pragmatic Bookshelf, 2013

\bibitem{ref:ANTLR4grammars} ANTLRv4 grammar repository, \url{https://github.com/antlr/grammars-v4}

\bibitem{ref:ECMAScript4MPS} ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}

\bibitem{ref:javascript} JavaScript ANTLRv4 grammar, \url{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}

>>>>

\bibitem{ref:ECMASCRIPT51} ECMAScript Language Specification, Edition 5.1 (June 2011), \url{http://www.ecma-international.org/ecma-262/5.1/Ecma-262.pdf}

@@

\bibitem{ref:JSON} The JSON Data Interchange Format, 1st Edition (October 2013), \url{http://www.ecma-international.org/publications/files/ECMA-ST/ECMA-404.pdf}

\bibitem{ref:ECMA4MPS} M. Lombardo. The ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}

@@
